New Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimisation Choices

### 2024-09-16 jso.qmd
minor corrections

  The projection pursuit (PP) guided tour interactively optimises a criterion function known as the PP index, to explore high-dimensional data by revealing interesting projections. Optimisation of some PP indexes can be non-trivial, if they are non-smooth functions, or the optima has a small "squint angle", detectable only from close proximity. To address these challenges, this study investigates the performance of a recently introduced swarm-based algorithm, Jellyfish Search Optimiser (JSO), for optimising PP indexes. The performance of JSO for visualising data is evaluated across various hyper-parameter settings and compared with existing optimisers. Additionally, methods for calculating the smoothness and squintability properties of the PP index are proposed. They are used to assess the optimiser performance in the presence of PP index complexities. A simulation study illustrates the use of these performance metrics to compare the JSO with existing optimisation methods available for the guided tour. The JSO algorithm has been implemented in the R package, `tourr`, and functions to calculate smoothness and squintability are available in the `ferrn` package. 

@@

Index functions are quite varied in form, partially depending on the data that is being projected. @fig-example-functions shows two examples. Huber plots [@huberplot] of 2D data sets are in (a) and (c), showing the PP index values for all 1D projections of the 2D data in polar coordinates, which reveals the form of these functions. The dashed circle is a baseline set at the average value, and the straight line marks the optimal projection. Plots (b) and (d) show the respective best projections of the data as histograms. Indexes like the holes, central mass and skewness [@cook1993projection] are generally smooth for most data sets, but capture only large patterns. Many indexes noisy and non-convex, requiring an effective and efficient optimisation procedure to explore the data landscape and achieve a globally optimal viewpoint of the data. The skewness index computed for trimodal data, in (a), is smooth with a large squint angle but has three modes, and thus is not convex. The binned normality index (a simple version of a non-normality index as described in @huber85) computed on the famous RANDU data, in (c), is noisier and has a very small squint angle. The discreteness cannot be seen unless the optimiser is very close to the optimal projection.

@@

Optimisation of PP is often discussed when new indexes are proposed [@posse95; @marie-sainte2010; @grochowski2011]. @cook1995grand tied the optimisation more closely to the index, when they introduced the PPGT, which monitors the optimisation visually so that the user can see the projected data leading in and out of the optima. An implementation is available in the `tourr` package [@tourr] in R [@R]. @RJ-2021-105 illustrated how to diagnose optimisation processes, particularly focusing on the guided tour, and revealed a need for improved optimisation. While improving the quality of the optimisation solutions in the tour is essential, it is also important to be able to view the data projections as the optimisation progresses. Integrating the guided tour with a global optimisation algorithm that is efficient in finding the global optimal and enables viewing of the projected data during the exploration process is a goal.

@@

The primary goal of the study reported here is to investigate the
performance of JSO in PP optimisation for the guided tour. It is of
interest to assess how quickly and closely the optimiser reaches a gobal
optima, for various PP indexes that may have differing complexities. To
observe the performance of JSO with different types of PP indexes,
metrics are introduced to capture specific properties of the index
including squintability (based on @barnett1981interpreting's squint
angle) and smoothness. Here, we mathematically define metrics for squintability and smoothness, which is a new contribution for PP research. A series of simulation experiments using various datasets and PP indexes are conducted to assess JSO's behaviour and its sensitivity to hyper-parameter choices (number of jellyfish and maximum number of tries). The relationship between the JSO performance, hyper-parameter choices and properties of PP indexes (smoothness and squintability) is analysed to provide guidance on selecting optimisers for practitioners using projection pursuit. Additionally, this work should guide the design of new PP indexes and facilitate better optimization for PP.

@@

- a pseudo-derivative, that searches locally for the best direction, based on differencing the index values for very close projections.
- a brute-force optimisation (CRS). 
- a modified brute force algorithm described in @posse95. 
- an essentially simulated annealing [@Bertsimas93] where the search space is reduced as the optimisation.
- a very localised search, to take tiny steps to get closer to the local maximum.

@@

There are numerous PP index functions available: introduced in @cook1993projection, @lee2005projection, @lee2010projection, @Grimm2016, @huber85, @Laa:2020wkm. Most are relatively simply defined, for any projection dimension, and implemented because they are relatively easy to optimise. A goal is to develop PP indexes based on scagnostics (@scag, @WW08), but the blockage is optimisation that these tend to be noisy, with potentially small squint angles.

An initial investigation of PP indexes, and the potential for
scagnostics is described in @laa_using_2020. To be useful here an
optimiser needs to be able to handle index functions which are possibly
not very smooth. In addition, because data structures might be
relatively fine, the optimiser needs to be able to find maxima that
occur with a small squint angle, that can only be seen from very close
by. One last aspect that is useful is for an optimiser to return local
maxima in addition to global because data can contain many different and
interesting features.

@@

```{=tex}
\begin{defn}
The Matérn covariance function $K$ is defined by
\begin{equation}
K(u)=K_{\nu,\eta,\ell}(u):=\eta^2\frac{\left(\sqrt{2\nu}\frac{u}{\ell}\right)^{\nu}}{\Gamma(\nu)2^{\nu-1}}\mathcal{K}_{\nu}\left(\sqrt{2\nu}\frac{u}{\ell}\right)\ , u\geq0\label{eq:matern}
\end{equation}
where $\nu>0$ is the smoothness parameter, $\eta$ is the outputscale, $\ell$ is the lengthscale, and $\mathcal{K}_\nu$ is
the modified Bessel function [DLMF 10.25]. A multivariate extension $K(u)$, $u\in\mathbb{R}^d$ can be obtained by products of univariate covariance functions \eqref{eq:matern}.
\end{defn}
```
The Matérn covariance function can be expressed analytically when $\nu$
is a half-integer, the most popular values in the literature being
$\frac{1}{2}$, $\frac{3}{2}$ and $\frac{5}{2}$ [@rasmussen2006gaussian]. The parameter $\nu$,
called *smoothness parameter*, controls the decay of the covariance
function. As such, it is an appropriate measure of smoothness of a
random field, as shown by the simulations on @fig-matern-1d and @fig-matern-2d. For example, @karvonen2023asymptotic showed that if a
function $f$ has a Sobolev index of $k$, then the smoothness parameter
estimate $\nu$ in \eqref{eq:matern} cannot be asymptotically less than
$k$. See the survey @porcu2024matern for additional results on the connection between the Matérn model and Sobolev spaces. An interesting result is that the asymptotic case
$\nu\rightarrow\infty$ coincides with the Gaussian kernel:
$K_\infty(u)=\exp(-u^{2}/2)$.

@@

In view of these results, the parameter $\nu$ is suggested as a measure
of the smoothness of the PP index function by fitting a Gaussian process
prior with Matérn covariance on a dataset generated by random
evaluations of the index function, as done at the initialization stage
of the jellyfish search optimization. There exist several R packages,
such as `GpGp` [@guinness2021gpgp] or `ExaGeoStatR` [@abdulah2023large],
to fit the hyperparameters of a GP covariance function on data, which is
usually done by maximum likelihood estimation. In this project, the
`GpGp` package is used.

```{=tex}
\begin{defn}
Let $\mathbf{A}=[A_1, \ldots, A_N] \in (\mathbf{R}^{p \times d})^N$ be d-dimensional projection bases, let $\mathbf{y}=[f(XA_1),\ldots,f(XA_N)]$ be the corresponding PP index values, and let $\mathbf{K}=[K_\theta(A_{i},A_{j})]_{1\leq i,j\leq N}\in\mathbb{R}^{N\times N}$ be the Matérn covariance matrix evaluated at the input bases, where the vector $\theta$ contains all the parameters of the multivariate Matérn covariance function $K$ (smoothness, outputscale, lengthscales). The log-likelihood of the parameters $\theta$ is defined by 
\begin{equation}
\mathcal{L}(\theta)=\log p(\mathbf{y}\left|\mathbf{A},\theta\right.)=-\frac{1}{2}\mathbf{y}^{\top}(\mathbf{K}+\sigma^{2}\mathbf{I})^{-1}\mathbf{y}-\frac{1}{2}\mathrm{\log}(\det(\mathbf{K}+\sigma^{2}\mathbf{I}))-\frac{N}{2}\log(2\pi)\, \label{eq:gp_log_likelihood}
\end{equation}
where the nugget parameter $\sigma$ is the standard deviation of the intrinsic noise of the Gaussian process.
The optimal parameters (including smoothness) are obtained by maximum log-likelihood
\begin{equation}
\theta^* = \underset{\theta}{\max}\mathcal{L}(\theta)
\end{equation}
The resulting optimal smoothness parameter $\nu$ is chosen as our smoothness metric.
\end{defn}
```

@@

Here the formal definition of projection distance and squint angle are given, before the definition of squintability. Two approaches to compute this metric numerically,
are described.

```{=tex}
\begin{defn}[projection distance]\label{def:proj-dist}
Recall that $A \in \mathbf{R}^{p \times d}$ is a $d$-dimensional orthonormal matrix. Let $A^*$ be the optimal matrix that achieves the maximum index value for a given data. The projection distance $d(A, A^*)$ between $A$ 
and $A^*$ is defined as 
$d(A, A^*) = \lVert AA^\prime - A^*A^{*\prime}\  \rVert _F$
where $\lVert . \rVert _F$ denotes the Frobenius norm, given by
$\lVert M \rVert _F = \sqrt{\sum_{ij} M_{ij}^2}$. 
\end{defn}
```

@@

\begin{equation}
\varsigma(f) = -c \times \max_{d} g'(d) \times \arg \max_{d} g'(d)
\label{eq-squintability}
\end{equation}

where $c$ is a constant scaling factor, $-\max_d g'(d)$ represents the 
largest gradient of $-g$ and $\arg \max_{d} g'(d)$ represents the projection distance at which this largest gradient is attained.

@@

It is expected that these two values should be both high in the case of high squintability (fast
increase in $f$ early on), and both low in the case of low squintability
(any substantial increase in $f$ happens very late, close to the optimal
angle). This suggests that their product \eqref{eq-squintability} should
provide a sensible measure of squintability. The multiplicative constant
$4$, which can be deemed arbitrary, does not change the interpretation
of the squintability metric $\varsigma$; it is here to adjust the range
of values of $\varsigma$ and simplify the explicit formula for
$\varsigma$ obtained later on.

@@

$$
\varsigma(g_1(d_1)) = -c  \max g'_1(d_1)  d_1 = -c  (-g_{\max}) d_1 \leq -c  (-g_{\max}) d_2 = -c \max g'_2(d_2)  d_2 = \varsigma(g_2(d_2))
$$

@@

From @barnett1981interpreting and @laa_using_2020, a large squint angle
implies that the objective function value is close to optimal even when
the perfect view to see the structure is far away. A small squint angle
means that the PP index value improves substantially only when the
perfect view is close by. As such, low squintability implies rapid
improvement in the index value when near the perfect view. For PP, a
small squint angle is considered to be undesirable because it means that
the optimiser needs to be very close to be able to "see" the optima.
Thus, it could be difficult for the optimiser to find the optima. <!--The
mathematical formulation of this intuition is proposed below:-->

It is expected that for a PP index with high squint angle, the
optimization (@eq-optimization) should make substantial progress early
on. Conversely, for a PP index with low squint angle, it might take a
long while for the optimization to make substantial progress, as the
candidate projections would need to be very close to the optimal one for
the structure of the index function to be visible enough to be amenable
to efficient optimization. This observation suggests that the extreme
values of $f'$ (the ones for which $f''=0$, assuming that $f$ is twice
differentiable), and the projection distances for which these values are
attained, are crucial in the mathematical definition of squintability.

@@

To compute the squintability metric \eqref{eq-squintability} in practice,
several approaches are possible. The first one is to propose a
parametric model for $f$, and use it to obtain an explicit formula for
$\varsigma$. Numerical experiments suggest a scaled sigmoid shape as
described below. Define

@@

$$
f(x)=(\theta_{1}-\theta_{4})\frac{\ell(x)-\ell(x_{\max})}{\ell(0)-\ell(x_{\max})}+\theta_{4}\ ,
$$ {#eq-parametric}

which depends on three additional parameters, $\theta_1$, $\theta_2$,
and $x_{\max}$, such that $f(0)=\theta_1$ and $f(x_{\max})=\theta_4$.
Under the parametric model (@eq-parametric), the squintability metric
\eqref{eq-squintability} can be shown to be equal to

@@

Alternatively, one can estimate \eqref{eq-squintability} in a nonparametric
way, for example by fitting $f$ using kernel regression, then
numerically estimate the angle at which $-f'$ attains its highest value.

@@

In the comparison of the JSO against the currently used CRS, as expected
the JSO vastly out-performs CRS, and provides a high probability of
finding the global optima. The JSO obtains the maximum more cleanly,
with a slightly higher index value, and plot of the projected data
showing the structure more clearly.

The JSO performance is affected by the hyper-parameters, with a higher
chance of reaching the global optima when more jellyfish are used and
the maximum number of tries is increased. However, it comes at a
computational cost, as expected. The performance declines if the
projection dimension increases and if the PP index has low
squintability. The higher the squintability the better chance the JSO
can find the optima. However, interestingly smoothness does not affect
the JSO performance.

>>>>

  The projection pursuit (PP) guided tour interactively optimises a criterion function known as the PP index, to explore high-dimensional data by revealing interesting projections. Optimisation of some PP indexes can be non-trivial, if they are non-smooth functions, or the optimum has a small "squint angle", detectable only from close proximity. To address these challenges, this study investigates the performance of a recently introduced swarm-based algorithm, Jellyfish Search Optimiser (JSO), for optimising PP indexes. The performance of JSO for visualising data is evaluated across various hyper-parameter settings and compared with existing optimisers. Additionally, methods for calculating the smoothness and squintability properties of the PP index are proposed. They are used to assess the optimiser performance in the presence of PP index complexities. A simulation study illustrates the use of these performance metrics to compare the JSO with existing optimisation methods available for the guided tour. The JSO algorithm has been implemented in the R package, `tourr`, and functions to calculate smoothness and squintability are available in the `ferrn` package. 

@@

Index functions are quite varied in form, partially depending on the data that is being projected. @fig-example-functions shows two examples. Huber plots [@huberplot] of 2D data sets are in (a) and (c), showing the PP index values for all 1D projections of the 2D data in polar coordinates, which reveals the form of these functions. The dashed circle is a baseline set at the average value, and the straight line marks the optimal projection. Plots (b) and (d) show the respective best projections of the data as histograms. Indexes like the holes, central mass and skewness [@cook1993projection] are generally smooth for most data sets, but capture only large patterns. Many indexes are noisy and non-convex, requiring an effective and efficient optimisation procedure to explore the data landscape and achieve a globally optimal viewpoint of the data. The skewness index computed for trimodal data, in (a), is smooth with a large squint angle but has three modes, and thus is not convex. The binned normality index (a simple version of a non-normality index as described in @huber85) computed on the famous RANDU data, in (c), is noisier and has a very small squint angle. The discreteness cannot be seen unless the optimiser is very close to the optimal projection.

@@

Optimisation of PP is often discussed when new indexes are proposed [@posse95; @marie-sainte2010; @grochowski2011]. @cook1995grand tied the optimisation more closely to the index, when they introduced the PPGT, which monitors the optimisation visually so that the user can see the projected data leading in and out of the optimum. An implementation is available in the `tourr` package [@tourr] in R [@R]. @RJ-2021-105 illustrated how to diagnose optimisation processes, particularly focusing on the guided tour, and revealed a need for improved optimisation. While improving the quality of the optimisation solutions in the tour is essential, it is also important to be able to view the data projections as the optimisation progresses. Integrating the guided tour with a global optimisation algorithm that is efficient in finding the global optimal and enables viewing of the projected data during the exploration process is a goal.

@@

The primary goal of the study reported here is to investigate the
performance of JSO in PP optimisation for the guided tour. It is of
interest to assess how quickly and closely the optimiser reaches a global
optimum, for various PP indexes that may have differing complexities. To
observe the performance of JSO with different types of PP indexes,
metrics are introduced to capture specific properties of the index
including squintability (based on @barnett1981interpreting's squint
angle) and smoothness. Here, we mathematically define metrics for squintability and smoothness, which is a new contribution for PP research. A series of simulation experiments using various datasets and PP indexes are conducted to assess JSO's behaviour and its sensitivity to hyper-parameter choices (number of jellyfish and maximum number of tries). The relationship between the JSO performance, hyper-parameter choices and properties of PP indexes (smoothness and squintability) is analysed to provide guidance on selecting optimisers for practitioners using projection pursuit. Additionally, this work should guide the design of new PP indexes and facilitate better optimization for PP.

@@

- a pseudo-derivative, that searches locally for the best direction, based on differencing the index values for very close projections.
- a brute-force optimisation (CRS). 
- a modified brute force algorithm described in @posse95. 
- an essentially simulated annealing [@Bertsimas93] where the search space is reduced during the optimisation.
- a very localised search, to take tiny steps to get closer to the local maximum.

@@

There are numerous PP index functions available: introduced in @huber85, @cook1993projection, @lee2005projection, @lee2010projection, @Grimm2016, @Laa:2020wkm. Most are relatively simply defined, for any projection dimension, and implemented because they are relatively easy to optimise. A goal is to develop PP indexes based on scagnostics (@scag, @WW08), but the blockage is their optimisation as these tend to be noisy, with potentially small squint angles.

An initial investigation of PP indexes, and the potential for
scagnostics is described in @laa_using_2020. To be useful here an
optimiser needs to be able to handle index functions that are possibly
not very smooth. In addition, because data structures might be
relatively fine, the optimiser needs to be able to find maxima that
occur with a small squint angle, that can only be seen from very close
by. One last aspect that is useful is for an optimiser to return local
maxima in addition to the global one because data can contain many different and
interesting features.

@@

```{=tex}
\begin{defn}
The Matérn covariance function $K$ is defined by
\begin{equation}
K(u)=K_{\nu,\eta,\ell}(u):=\eta^2\frac{\left(\sqrt{2\nu}\frac{\left\Vert u\right\Vert}{\ell}\right)^{\nu}}{\Gamma(\nu)2^{\nu-1}}\mathcal{K}_{\nu}\left(\sqrt{2\nu}\frac{\left\Vert u\right\Vert}{\ell}\right)\ ,\label{eq:matern}
\end{equation}
where $\left\Vert u\right\Vert$ is the Euclidean norm of $u\in\mathbb{R}^{p{\times}d}$, $\nu>0$ is the smoothness parameter, $\eta$ is the outputscale, $\ell$ is the lengthscale, and $\mathcal{K}_\nu$ is
the modified Bessel function [DLMF 10.25].
\end{defn}
```
The Matérn covariance function can be expressed analytically when $\nu$
is a half-integer, the most popular values in the literature being
$\frac{1}{2}$, $\frac{3}{2}$ and $\frac{5}{2}$ [@rasmussen2006gaussian]. The parameter $\nu$,
called *smoothness parameter*, controls the decay of the covariance
function. As such, it is an appropriate measure of smoothness of a
random field, as shown by the simulations on @fig-matern-1d and @fig-matern-2d. For example, @karvonen2023asymptotic showed that if a
function $f$ has a Sobolev index of $k$, then the smoothness parameter
estimate $\nu$ in \eqref{eq:matern} cannot be asymptotically less than
$k$. See the survey @porcu2024matern for additional results on the connection between the Matérn model and Sobolev spaces. An interesting result is that the asymptotic case
$\nu\rightarrow\infty$ coincides with the Gaussian kernel:
$K_\infty(u)=\exp(-{\left\Vert u\right\Vert}^{2}/2)$.

@@

In view of these results, the parameter $\nu$ is suggested as a measure
of the smoothness of the PP index function by fitting a Gaussian process
prior with Matérn covariance on a dataset generated by random
evaluations of the index function, as done at the initialization stage
of the jellyfish search optimization. There exist several R packages,
such as `GpGp` [@guinness2021gpgp] or `ExaGeoStatR` [@abdulah2023large],
to fit the hyper-parameters of a GP covariance function on data, which is
usually done by maximum likelihood estimation. In this project, the
`GpGp` package is used.

```{=tex}
\begin{defn}
Let $\mathbf{A}=[A_1, \ldots, A_N] \in (\mathbf{R}^{p \times d})^N$ be d-dimensional projection bases, let $\mathbf{y}=[f(XA_1),\ldots,f(XA_N)]$ be the corresponding PP index values, and let $\mathbf{K}=[K_\theta(A_{i}-A_{j})]_{1\leq i,j\leq N}\in\mathbb{R}^{N\times N}$ be the Matérn covariance matrix evaluated at the input bases, where the vector $\theta$ contains all the parameters of the multivariate Matérn covariance function $K$ (smoothness, outputscale, lengthscales). The log-likelihood of the parameters $\theta$ is defined by 
\begin{equation}
\mathcal{L}(\theta)=\log p(\mathbf{y}\left|\mathbf{A},\theta\right.)=-\frac{1}{2}\mathbf{y}^{\top}(\mathbf{K}+\sigma^{2}\mathbf{I})^{-1}\mathbf{y}-\frac{1}{2}\mathrm{\log}(\det(\mathbf{K}+\sigma^{2}\mathbf{I}))-\frac{N}{2}\log(2\pi)\, \label{eq:gp_log_likelihood}
\end{equation}
where the nugget parameter $\sigma$ is the standard deviation of the intrinsic noise of the Gaussian process.
The optimal parameters (including smoothness) are obtained by maximum log-likelihood
\begin{equation}
\theta^* = \underset{\theta}{\max}\mathcal{L}(\theta)
\end{equation}
The resulting optimal smoothness parameter $\nu$ is chosen as our smoothness metric.
\end{defn}
```

@@

Here the formal definitions of projection distance and squint angle are given, before the definition of squintability. Two approaches to compute this metric numerically are described.

```{=tex}
\begin{defn}[projection distance]\label{def:proj-dist}
Recall that $A \in \mathbf{R}^{p \times d}$ is a $d$-dimensional orthonormal matrix. Let $A^*$ be the optimal matrix that achieves the maximum index value for a given data. The projection distance $d(A, A^*)$ between $A$ 
and $A^*$ is defined as 
$d(A, A^*) = \lVert AA^\prime - A^*A^{*\prime}\,\rVert _F$
where $\lVert . \rVert _F$ denotes the Frobenius norm, given by
$\lVert M \rVert _F = \sqrt{\sum_{ij} M_{ij}^2}$. 
\end{defn}
```

@@

\begin{equation}
\varsigma(f) = -4 \times \max_{d} g'(d) \times \arg \max_{d} g'(d)
\label{eq-squintability}
\end{equation}

where $4$ is a constant scaling factor, $-\max_d g'(d)$ represents the 
largest gradient of $-g$ and $\arg \max_{d} g'(d)$ represents the projection distance at which this largest gradient is attained.

@@

It is expected that these two values should be both high in the case of high squintability (fast
increase in $g$ early on), and both low in the case of low squintability
(any substantial increase in $g$ happens very late, close to the optimal
angle). This suggests that their product \eqref{eq-squintability} should
provide a sensible measure of squintability. The multiplicative constant
$4$, which can be deemed arbitrary, does not change the interpretation
of the squintability metric $\varsigma$; it is here to adjust the range
of values of $\varsigma$ and simplify the explicit formula for
$\varsigma$ obtained later on.

@@

$$
\varsigma(g_1(d_1)) = -4  \max g'_1(d_1)  d_1 = -4  (-g_{\max}) d_1 \leq -4  (-g_{\max}) d_2 = -4 \max g'_2(d_2)  d_2 = \varsigma(g_2(d_2))
$$

@@

From @barnett1981interpreting and @laa_using_2020, a large squint angle
implies that the objective function value is close to optimal even when
the perfect view to see the structure is far away. A small squint angle
means that the PP index value improves substantially only when the
perfect view is close by. As such, low squintability implies rapid
improvement in the index value when near the perfect view. For PP, a
small squint angle is considered to be undesirable because it means that
the optimiser needs to be very close to be able to "see" the optimum.
Thus, it could be difficult for the optimiser to find the optimum. <!--The
mathematical formulation of this intuition is proposed below:-->

It is expected that for a PP index with high squint angle, the
optimization (@eq-optimization) should make substantial progress early
on. Conversely, for a PP index with low squint angle, it might take a
long while for the optimization to make substantial progress, as the
candidate projections would need to be very close to the optimal one for
the structure of the index function to be visible enough to be amenable
to efficient optimization. This observation suggests that the extreme
values of $g'$ (the ones for which $g''=0$, assuming that $g$ is twice
differentiable), and the projection distances for which these values are
attained, are crucial in the mathematical definition of squintability.

@@

To compute the squintability metric \eqref{eq-squintability} in practice,
several approaches are possible. The first one is to propose a
parametric model for $g$, and use it to obtain an explicit formula for
$\varsigma$. Numerical experiments suggest a scaled sigmoid shape as
described below. Define

@@

$$
g(x)=(\theta_{1}-\theta_{4})\frac{\ell(x)-\ell(x_{\max})}{\ell(0)-\ell(x_{\max})}+\theta_{4}\ ,
$$ {#eq-parametric}

which depends on three additional parameters, $\theta_1$, $\theta_2$,
and $x_{\max}$, such that $g(0)=\theta_1$ and $g(x_{\max})=\theta_4$.
Under the parametric model (@eq-parametric), the squintability metric
\eqref{eq-squintability} can be shown to be equal to

@@

Alternatively, one can estimate \eqref{eq-squintability} in a nonparametric
way, for example by fitting $g$ using kernel regression, then
numerically estimate the angle at which $-g'$ attains its highest value.

@@

In the comparison of the JSO against the currently used CRS, as expected
the JSO vastly outperforms CRS, and provides a high probability of
finding the global optimum. The JSO obtains the maximum more cleanly,
with a slightly higher index value, and plot of the projected data
showing the structure more clearly.

The JSO performance is affected by the hyper-parameters, with a higher
chance of reaching the global optimum when more jellyfish are used and
the maximum number of tries is increased. However, it comes at a
computational cost, as expected. The performance declines if the
projection dimension increases and if the PP index has low
squintability. The higher the squintability the better chance the JSO
can find the optimum. However, interestingly smoothness does not affect
the JSO performance.
